Theories of deliberated judgment

### 2022-08-07 simple.tex
Defining DC instead of capturing it; gave up observing incompleteness

While $I$ and $\allargs$ are fundamental sets whose elements are considered elementary, a few notations and concepts are required to define ${\allleadsto}$.
Throughout the article, $\N$ includes zero. %and $\N^* = \N \setminus \set{0}$. 
Given $j, l \in \N$, the notation $\intvl{j, l}$ represents $\set{k \in \N \suchthat j ≤ k ≤ l}$ (thus $\forall j \in \N: \intvl{j, 0} = \emptyset$).
Define $\allhist = \bigcup_{k \in \N} \allargs^{\intvl{1, k}}$ as the finite sequences of arguments, including the empty sequence. 
A generic element of $\allhist$ (a sequence of arguments) is denoted by $\hist$ and called an argumentative path, or simply a path when this raises no ambiguity. Its length is denoted by $\card{\hist} \in \N$.
Thus, given an argumentative path $\hist \in \allhist$, $\forall k \in \intvl{1, \card{\hist}}$, $\hist_k \in \allargs$, and $\hist = \emptyset ⇔ \card{\hist} = 0$.

@@

\subsection{Deliberated choice}
The deliberated choice of $i$ given the exposure protocol $(I, \allargs, {\allleadsto})$ is denoted $p_i \in P$, with the semantics that $p_i$ represents the preference of $i$ over the proposed set of two acts, after deliberation, given the exposure protocol. In other words, if $p_i = \phi$, the individual considers the first act strictly better than the second one, after deliberation (and conversely for $p_i = ¬\phi$); and if $p_i = 0$, the deliberated choice of $i$ considers neither no act as strictly better than the alternative. This latter possibility covers both the situation of indifference between both options (in the sense of being evaluated by $i$ as being exactly equally worth), which will be further developed in \cref{sec:multiple}, or the situation that no act is deliberately strictly better than the other one and that they are also not deliberately exactly equivalent to each other.
\footnote{It should be clear that $P$ as used as the codomain of ${\leadsto}$ and as the set of possible values of $p_i$ denote different entities. The letter is reused simply to reduce notations. The value ${\leadsto}(i, \alpha) \in P$ designates the directly observed choice of $i$ whereas the value $p_i \in P$ designates the non directly observable deliberated choice of $i$%; furthermore, $0$ in particular may acquire a very different meaning as the value ${\leadsto}(i, \alpha)$ than as the value $p_i = 0$
.}

This latter possibility may look unpleasant, but it is a fact that facing some difficult problems, individuals sometimes are undecided, and it is an a priori empirical possibility that the individual remains undecided (or even, becomes \emph{more} undecided) when considering multiple arguments. That classical protocols of natural choice do not allow to observe such incompleteness has often been considered a weakness of those protocols, that our richer observational basis permits to overcome.

The formal definition of $p_i$ given $(I, \allargs, {\allleadsto})$ requires to define the notion of decisive argument.

@@

Given $i \in I, p \in P, \ar \in \allargs$, define $\ar \ileadstost p$ iff $\forall \hist \in \allhist \suchthat \ar \in \histend: \hist \ileadsto p$.
When $\ar \ileadstost p$, $\ar$ is said to be a decisive argument for $\prop$ from $i$’s point of view given the exposure protocol. 
The deliberated judgment of $i$ is then considered as the set of propositions supported by decisive arguments.
\begin{axiom}[Decisive arguments define deliberated judgments]
	\label{ax:decisive}
	$\forall i \in I, p \in P: 
		p \in P_i ⇔ [\exists \ar \in \allargs \suchthat \ar \ileadstost \prop].$
\end{axiom}

@@

\subsection{Theories using a static decisive argument}
\label{sec:static}
\Cref{ex:tutor} suggests a simple form for theories of deliberated choice: suffice to determine an argument $\ar \in \allargs$ and an alternative $\prop \in \Phi$ such that the theory claims that $\forall i \in I: \ar \ileadstost \prop$ and that $\prop \in \iprops$.

\begin{definition}[Static theory]
	\label{def:static}
	A static theory is a pair $(\ar, \prop)$ for some $\ar \in \allargs$ and some $\prop \in \Phi$.
\end{definition}
Such a theory is said to be truthful iff $\forall i \in I: \ar \ileadstost \prop$, and to be correct iff $\prop \in \iprops$.
We can then say that such theories are valid, meaning that their truthfulness implies their correctness. Let us make this (very simple) result stand out as a theorem, because a parallel will be usefully made with more complicated theories.
%Note that this result immediately follows from \cref{ax:decisive}.
\begin{theorem}[Static theories are valid]
	Any truthful static theory is correct.
\end{theorem}
\begin{proof}
	The result immediately follows from \cref{ax:decisive}.
\end{proof}
Importantly, truthfulness can indeed be checked using our observational basis, that is, if $\exists i \in I \suchthat \ar \nileadstost \prop$, then it is possible to exhibit this fact experimentally.

@@

The following example illustrates this same point in a simpler setting which permits a complete definition of $\allargs$ and $\leadsto$.%, which permits defining the setting completely at the price of a reduced realism.
\begin{example}[A decision problem about two lotteries]
	\label{ex:lichtenstein}
	In a famous article about preference reversals, \citet{lichtenstein_reversals_2006} discuss an example involving two lotteries. The lottery $P$ lets the subject win four dollars with 99\% chance and lose ten cents with 1\% chance; in summary, $P = (.99, +\$4; .01, −\$0.1)$. The lottery $D$ is, using the same notations: $(.33, +\$16; .67, −\$2)$. The letter $P$ stands for probability: one salient aspect of $P$ is the very high chance of winning. The letter $D$ stands for dollar: one salient aspect of $D$ is the high gain it gives (compared to $P$) in dollar amount if the subject is lucky.

@@

\subsection{Argumentative theories of deliberated choice}
\label{sec:dynamic}
An argumentative theory of \ac{DC} is defined as a tuple $(\gleadsto, \dg)$, where ${\gleadsto} \subseteq \allargs × \allprops$ is an initial argumentation relation and $\dg$ is a defense relation; satisfying bounded depth.

>>>>

While $I$ and $\allargs$ are fundamental sets whose elements are considered elementary, a few notations and concepts are required to define ${\allleadsto}$.
Throughout the article, $\N$ includes zero. %and $\N^* = \N \setminus \set{0}$. 
Given $j, l \in \N$, the notation $\intvl{j, l}$ represents $\set{k \in \N \suchthat j ≤ k ≤ l}$ (thus $\forall j \in \N: \intvl{j, 0} = \emptyset$).
Define $\allhist = \bigcup_{k \in \N} \allargs^{\intvl{1, k}}$ as the finite sequences of arguments, including the empty sequence. 
A generic element of $\allhist$ (a sequence of arguments) is denoted by $\hist$ and called an argumentative path, or simply a path when this raises no ambiguity. Its length is denoted by $\card{\hist} \in \N$.
Thus, given an argumentative path $\hist \in \allhist$, $\forall k \in \intvl{1, \card{\hist}}: \hist_k \in \allargs$, and $\hist = \emptyset ⇔ \card{\hist} = 0$.

@@

\subsection{Deliberated choice}
The deliberated choice of $i$ given the exposure protocol $(I, \allargs, {\allleadsto})$ is denoted $P_i \subset P$.
Its formal defnition requires the notion of decisive argument.

@@

Given $i \in I, p \in P, \ar \in \allargs$, define $\ar \ileadstost p$ iff $\forall \hist \in \allhist \suchthat \ar \in \histend: \hist \ileadsto p$.
When $\ar \ileadstost p$, $\ar$ is said to be a decisive argument for $\prop$ from $i$’s point of view given the exposure protocol.

The deliberated choice of $i$ is then considered as the set of propositions supported by decisive arguments.
\begin{definition}[Deliberated choice]
	\label{def:decisive}
	$\forall i \in I, p \in P: 
		p \in P_i ⇔ [\exists \ar \in \allargs \suchthat \ar \ileadstost p].$
\end{definition}

It follows from the consideration of “end” as the last \emph{two} arguments that decisive arguments can argue for at most one proposition, thus, that the deliberated choice cannot hold more than one proposition.
\begin{theorem}[Deliberated choice singularity]
	$\forall i \in I: \card{P_i} ≤ 1$.
\end{theorem}
\begin{proof}
	Assume that $\exists \ar, \ar' \in \allargs, p, p' \in P \suchthat \ar \ileadstost p \land \ar' \ileadstost p'$.
	Since $\ar \ileadstost p$, $(\ar, \ar') \ileadsto p$, and since $\ar' \ileadstost p'$, $(\ar, \ar') \ileadsto p'$, hence $p' = p$.
\end{proof}

The intended semantics is that $P_i$ represents the preference of $i$ over the proposed set of two acts, after deliberation, given the exposure protocol. In other words, if $P_i = \set{\phi}$, the individual considers the first act strictly better than the second one, after deliberation (and conversely for $P_i = \set{¬\phi}$); if $P_i = \set{0}$, the deliberated judgment of $i$ is that both acts are equally worth; and if $P_i = \emptyset$, no act is deliberately strictly better than the other one and they are also not deliberately exactly equivalent to each other.

This latter possibility may look unpleasant, but it is a fact that facing some difficult problems, individuals sometimes are undecided, and it is an a priori empirical possibility that the individual remains undecided (or even, becomes \emph{more} undecided) when considering multiple arguments.

@@

\subsection{Theories using a static decisive argument}
\label{sec:static}
\Cref{ex:tutor} suggests a simple form for theories of deliberated choice: suffice to determine an argument $\ar \in \allargs$ and an alternative $\prop \in \Phi$ such that the theory claims that $\forall i \in I: \ar \ileadstost \prop$ and that $\forall i \in I: \prop \in \iprops$.

\begin{definition}[Static theory]
	\label{def:static}
	A static theory is a pair $(\ar, \prop)$ for some $\ar \in \allargs$ and some $\prop \in \Phi$.
\end{definition}
Such a theory is said to be truthful iff $\forall i \in I: \ar \ileadstost \prop$, and to be correct iff $\forall i \in I: \prop \in \iprops$.
We can then say that such theories are valid, meaning that their truthfulness implies their correctness. Let us make this (very simple) result stand out as a theorem, because a parallel will be usefully made with more complicated theories.
%Note that this result immediately follows from \cref{ax:decisive}.
\begin{theorem}[Static theories are valid]
	Any truthful static theory is correct.
\end{theorem}
\begin{proof}
	This follows immediately from \cref{def:decisive}.
\end{proof}
Importantly, truthfulness can indeed be checked using our observational basis, that is, if $\exists i \in I \suchthat \ar \nileadstost \prop$, then it is possible to exhibit this fact experimentally.

@@

The following example illustrates this same point in a constrained setting which permits a complete definition of $\allargs$ and $\leadsto$.%, which permits defining the setting completely at the price of a reduced realism.
\begin{example}[A decision problem about two lotteries]
	\label{ex:lichtenstein}
	In a famous article about preference reversals, \citet{lichtenstein_reversals_2006} discuss an example involving two lotteries. The lottery $P$ lets the subject win four dollars with 99\% chance and lose ten cents with 1\% chance; in summary, $P = (.99, +\$4; .01, −\$0.1)$. The lottery $D$ is, using the same notations: $(.33, +\$16; .67, −\$2)$. The letter $P$ stands for probability: one salient aspect of $P$ is the very high chance of winning. The letter $D$ stands for dollar: one salient aspect of $D$ is the high gain it gives (compared to $P$) in dollar amount if the subject is lucky.

@@

\subsection{Argumentative theories of deliberated choice}
\label{sec:dynamic}
An argumentative theory of \ac{DC} is defined as a tuple $(p, \gleadsto, \dg)$, where $p \in P$ is the proposition supported by the theory, ${\gleadsto} \subseteq \allargs × \allprops$ is an initial argumentation relation and $\dg$ is a defense relation; satisfying bounded depth.
